---
title: "Using the Benford's law to assess the quality of COVID-19 reported data in Vietnam"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo       = TRUE,
                      message    = FALSE,
                      fig.retina = 2,
                      fig.align  = "center")
```

## Preambule

Analysis following [previously published work](https://osf.io/74xjc) on
Brazilian data, using the [`benford.analysis`](https://cran.r-project.org/web/packages/benford.analysis/index.html)
R package.

## Conclusions

To be completed.

## Packages and global variables

```{r}
clean_folder <- "~/Dropbox/aaa/studies/oucomo/clean data/"
```

```{r}
library(readxl)
library(magrittr)
library(dplyr)
library(tidyr)
library(purrr)
library(benford.analysis)
library(tibble)
library(lubridate)
library(sf)
library(stringr)
library(stringi)
```

## Utilitary functions

A function to extract the statistics of a Benford analysis we want to look at:

```{r}
extract_statistics <- function(x) {
  with(x, c(with(info, c(n, n.second.order)),
            stats$chisq$statistic,
            MAD,
            MAD.conformity,
            distortion.factor,
            stats$mantissa.arc.test$statistic))
}
```

A function to make a table of statistics (in column) for each of the provinces
(in row):

```{r}
make_statistics_table <- function(x) {
  x %>% 
    map_df(extract_statistics) %>% 
    t() %>% 
    as.data.frame() %>% 
    rownames_to_column() %>% 
    setNames(c("province", "n", "n2", "Chisq", "MAD", "MAD conformity", "DF", "Mantissa")) %>% 
    as_tibble() %>% 
    mutate_at(vars(starts_with("n")), as.integer) %>% 
    mutate_at(vars(MAD, DF, Chisq, Mantissa), as.numeric)
}
```

A function that plots the diagnostic plots for each of the provinces:

```{r}
plot_digits_distribution <- function(x, y) {
  xmarks <- barplot(x[["bfd"]]$data.dist.freq, col = 4,
                    xlab = "digits", ylab = "frequency", 
                    ylim = c(0, 1.1 * max(c(x[["bfd"]]$data.dist.freq,
                                            x[["bfd"]]$benford.dist.freq))))
  axis(1, at = xmarks, labels = x[["bfd"]]$digits)
  lines(xmarks, x[["bfd"]]$benford.dist.freq, lwd = 2, col = 2)
  title(y, line = -1)
}
```

## Data

### Vietnam 2019 census data

The 2019 census data (`Vinh Long` is missing):

```{r}
census <- paste0(clean_folder, "census2019.rds") %>%
  readRDS() %>% 
  group_by(province) %>% 
  summarise(popsize = sum(n)) %>% 
  mutate_at("province", stri_trans_general, "Latin-ASCII") %>% 
  mutate_at("province", str_remove, "Thanh pho |Tinh ") %>% 
  bind_rows(tibble(province = "Vinh Long", popsize = 1141677))
```

### Geographic data of Vietnam

Downloading the file if not in the folder:

```{r}
if(!file.exists("gadm36_VNM_1_sf.rds"))
  download.file("https://biogeo.ucdavis.edu/data/gadm3.6/Rsf/gadm36_VNM_1_sf.rds",
                "gadm36_VNM_1_sf.rds")
```

Loading the data:

```{r}
gadm <- "gadm36_VNM_1_sf.rds" %>%
  readRDS() %>% 
  st_set_crs(4326) %>% 
  transmute(province = VARNAME_1, geometry) %>% 
  mutate(centroid    = map(geometry, st_centroid),
         coordinates = map(centroid, st_coordinates),
         latitude    = map(coordinates, extract2, 2)) %>% 
  left_join(census, "province")
```

### Rogier's COVID-19 data

Reading the data:

```{r}
rogier <-  paste0(clean_folder, "Fourth cluster First wave.xlsx") %>% 
  read_excel() %>% 
  rename(date = `...1`)
```

Getting rid off whatever is below the `Total` row:

```{r}
rogier %<>% head(which(rogier$date == "Total") - 1)
```

Reformatting the data:

```{r}
rogier %<>% 
  select(date, `An Giang`:last_col()) %>% 
  filter(! is.na(date)) %>% 
  mutate_all(replace_na, 0) %>% 
  mutate_all(as.integer) %>% 
  mutate_at("date", as.Date, origin = "1899-12-30")
```

which looks like:

```{r}
rogier
```

For compatibility of other data sets, we want to rename `Hue` into
`Thua Thien Hue`:

```{r}
names(rogier) <- str_replace(names(rogier), "Hue", "Thua Thien Hue")
```

### NCSC's COVID-19 data

Data automatically collected from [NCSC](https://covid.ncsc.gov.vn):

```{r}
ncsc <- readRDS(paste0(clean_folder, "NCSC data/covid.rds")) %>% 
  filter(var == "incidence") %>% 
  select(-var) %>% 
  mutate_at("province", str_replace, "TP HCM", "Ho Chi Minh")
```

which looks like this:

```{r}
ncsc
```

### Comparing Rogier's and NCSC data

A long format of Rogier's data:

```{r}
rogier_long <- pivot_longer(rogier, -date, "province", values_to = "n_rogier")
```

Merging Rogier's and NCSC's data:

```{r}
two_sources <- rogier_long %>% 
  full_join(ncsc, c("date", "province")) %>% 
  mutate_at(vars(starts_with("n")), replace_na, 0) %>% 
  filter(n_rogier > 0 | n > 0)
```

Plots to compare the two sources of data:

```{r fig.width = 2 * 4.24725, fig.height = 1.2 * 10 * 3.5 / 16}
par(mfrow = c(1, 3), plt = c(.15, .9, .17, .9))

two_sources %$% 
  plot(n_rogier, n, xlab = "Rogier's data", ylab = "NCSC data",
       col = adjustcolor(4, .5))
abline(0, 1)

two_sources %$% 
  plot(n_rogier, n, xlab = "Rogier's data", ylab = "NCSC data",
       col = adjustcolor(4, .5), xlim = c(0, 9000), ylim = c(0, 9000))
abline(0, 1)

two_sources %>% 
  mutate_at(vars(starts_with("n")), ~ .x + .1) %$% 
  plot(n_rogier, n, xlab = "Rogier's data", ylab = "NCSC data",
       col = adjustcolor(4, .5), log = "xy")
abline(0, 1)
```

Comparing the dates of zero-counts for the two data sets:

```{r}
two_sources %>% 
  filter(n < 1) %>% 
  pull(date) %>% 
  plot(., jitter(rep(1, length(.))), axes = FALSE, col = 2, ann = FALSE,
       xlim = ymd(c(20200101, 20211231)), ylim = c(.8, 1.1))

two_sources %>% 
  filter(n_rogier < 1) %>% 
  pull(date) %>% 
  points(., jitter(rep(.9, length(.))), col = 4)

month_val <- seq(1, 12, 2)
axis(1,
     ydm(paste0("2020-1-", month_val), paste0("2021-1-", month_val), "2022-1-1"),
     c(paste0(month.abb[month_val], "-20"), paste0(month.abb[month_val], "-21"), "Jan-22"))

month_val2 <- 1:12
abline(v = ydm(paste0("2020-1-", month_val2), paste0("2021-1-", month_val2), "2022-1-1")[-c(1, 2)])

text(ymd(20200115), 1, "NCSC", col = 2)
text(ymd(20200115), .9, "Rogier", col = 4)
```

The distribution of the counts in one data set when it's 0 in the other data set:

```{r fig.width = 2 * 4.24725, fig.height = 1.5 * 10 * 3.5 / 16}
par(mfrow = c(1, 2), plt = c(.15, .95, .17, .95))

hist2 <- function(..., color) {
  hist(..., n = 1000, col = color, border = color, yaxs = "i",
       main = NA, ylab = "frequency")
}

two_sources %>% 
  filter(n_rogier < 1) %>% 
  pull(n) %>% 
  hist2(color = 4, xlab = "value in NCSC when 0 count in Rogier")

two_sources %>% 
  filter(n < 1) %>% 
  pull(n_rogier) %>% 
  hist2(color = 2, xlab = "value in Rogier when 0 count in NCSC")
```

A function to plot the time series from the 2 sources for 1 province:

```{r}
plot_prov_ts <- function(prov, xlim = ymd(c(20200101, 20220101)), lwd = 1) {
  ncsc %>% 
    filter(province == prov) %$%
    plot(date, n, type = "l", col = 2, xlab = NA, ylab = "incidence",
         xlim = xlim, lwd = lwd)

  rogier_long %>% 
    filter(province == prov) %$%
    lines(date, n_rogier, col = 4, lty = 2, lwd = lwd)
}
```

Comparing the time series of the two data sets for Ho Chi Minh city:

```{r}
plot_prov_ts("Ho Chi Minh", ymd(c(20210601, 20220101)), lwd = 2)
legend("topright", legend = c("NCSC", "Rogier"),
       col = c(2, 4), lty = 1:2, bty = "n", lwd = 2)
```

Comparing for all the provinces:

```{r fig.width = 2 * 4.24725, fig.height = 20 * 3.5}
par(mfrow = c(32, 2), plt = c(.2, .97, .17, .9))

ncsc %>% 
  pull(province) %>% 
  unique() %>% 
  sort() %>% 
  walk(~ {plot_prov_ts(.x); title(.x, line = -1)})
```

## Analyzing Rogier's data

### All the data

Removing the dates and filtering out the zeros:

```{r}
rogier1 <- rogier %>% 
  select(-date) %>% 
  map(~ .x[.x > 0])
```

Performing the Benford analysis on each of the 63 provinces:

```{r}
benford_analyses_rogier1 <- map(rogier1, benford, number.of.digits = 1)
```

Extracting all these statistics for each of the 63 provinces. Interpretation is
as follows:

* `n`: number of data points on which the statistics are calculated.
* `n2`: number of data points on which the second order test is performed.
* `Chisq`: classical chi-square statistic. Note though that it is highly
           sensitive to the sample size and tends to reject the null even for
           small departures from the expected distribution. `MAD` statistic is
           thus to be preferred.
* `MAD`: Mean Absolute Deviation. The MAD test is more robust since it ignores
         the number of records. The higher the MAD, the larger the average
         difference between the observed and theoretical distributions. MAD
         values above 0.015 suggest nonconformity.
* `MAD conformity`: interpretation of the `MAD` statistic value.
* `DF`: Distortion Factor. The DF statistic examines the digit patterns to
        indicate whether the data appear to be underestimated or overestimated
        and the deformity’s magnitude.
* `Mantissa`: expected to be 0.5. Values less than 0.5 suggest that figures tend
              to be manipulated downward whereas it's the opposite for values
              higher than 0.5.

```{r}
table_rogier1 <- make_statistics_table(benford_analyses_rogier1)
print(table_rogier1, n = Inf)
```

Plotting the diagnostic plots for each of the 63 provinces:

```{r fig.width = 2 * 4.24725, fig.height = 10 * 3.5}
par(mfrow = c(16, 4), plt = c(.2, .97, .17, .9))
walk2(benford_analyses_rogier1,
      names(benford_analyses_rogier1),
      plot_digits_distribution)
```

### Data from January 2021

Same analysis as above but with data from January 2021 only:

```{r fig.width = 2 * 4.24725, fig.height = 10 * 3.5}
rogier2 <- rogier %>% 
  filter(date > ymd(20210101)) %>% 
  select(-date) %>% 
  map(~ .x[.x > 0])

benford_analyses_rogier2 <- map(rogier2, benford, number.of.digits = 1)

table_rogier2 <- make_statistics_table(benford_analyses_rogier2)
print(table_rogier2, n = Inf)

par(mfrow = c(16, 4), plt = c(.2, .97, .17, .9))
walk2(benford_analyses_rogier2,
      names(benford_analyses_rogier2),
      plot_digits_distribution)
```

Let's compare with the table we get with all the data:

```{r fig.width = 2 * 4.24725, fig.height = 2 * 3.5}
vars <- setdiff(names(table_rogier1), c("province", "MAD conformity"))
par(mfrow = c(2, 3), plt = c(.2, .97, .17, .9), cex = 1)
walk(vars, ~ {
  plot(table_rogier1[[.x]], table_rogier2[[.x]], col = 4,
       xlab = "all the data", ylab = "from Jan 2021")
  abline(0, 1)
  title(.x, line = -1)})
```

### Specific comparisons

#### Group 1: Hai Duong and Quang Ninh between 1 January 2021 and 1 April 2021

```{r}
group1 <- rogier %>% 
  transmute(date, n = `Hai Duong` + `Quang Ninh`) %>% 
  filter(ymd(20210101) < date, date < ymd(20210401)) %>% 
  select(-date) %>% 
  map(~ .x[.x > 0])
```

#### Group 2: Bac Giang and Bac Ninh between 1 April 2021 and 1 August 2021

```{r}
group2 <- rogier %>% 
  transmute(date, n = `Bac Giang` + `Bac Ninh`) %>% 
  filter(ymd(20210401) < date, date < ymd(20210801)) %>% 
  select(-date) %>% 
  map(~ .x[.x > 0])
```

#### Group 3: HCMC and Mekong between 1 May 2021 and 15 October 2021

The fourth wave:

```{r}
wave4 <- rogier %>% 
  filter(ymd(20210501) < date, date < ymd(20211015)) %>% 
  select(-date)
```

Merging with GADM data:

```{r}
wave4gadm <- wave4 %>% 
  colSums() %>% 
  list() %>% 
  data.frame() %>% 
  setNames("n") %>% 
  rownames_to_column("province") %>% 
  left_join(gadm, ., "province") %>% 
  mutate(incidence = 100000 * n / popsize)
```

The south wave in the south:

```{r}
wave4mekong <- wave4gadm %>% 
  filter(latitude < 12, incidence > 50)
```

Which looks like this (in red):

```{r}
par(plt = c(0, 1, 0, 1))

gadm %>% 
  st_geometry() %>% 
  plot(col = "grey")

wave4mekong %>% 
  st_geometry() %>% 
  plot(add = TRUE, col = "red")
```

Group 3 is then:

```{r}
group3 <- wave4 %>% 
  magrittr::extract(wave4mekong$province) %>% 
  rowSums() %>% 
  list()
```

#### Group 4: the whole country after 15 October 2021

```{r}
group4 <- rogier %>% 
  filter(date > ymd(20211015)) %>% 
  select(-date) %>% 
  rowSums() %>% 
  list()
```

Let's group the 4 groups into a list:

```{r}
groups <- c(group1, group2, group3, group4) %>% 
  setNames(paste0("group", 1:4))
```

on which we can rerun the previous analyses:

```{r fig.width = 2 * 4.24725, fig.height = 10 * 3.5 / 16}
benford_analyses_rogier_4groups <- map(groups, benford, number.of.digits = 1)

make_statistics_table(benford_analyses_rogier_4groups)

par(mfrow = c(1, 4), plt = c(.2, .97, .17, .9))
walk2(benford_analyses_rogier_4groups,
      names(benford_analyses_rogier_4groups),
      plot_digits_distribution)
```

## Analyzing NCSC's data

### All the data

```{r fig.width = 2 * 4.24725, fig.height = 10 * 3.5}
ncsc_wide <- pivot_wider(ncsc, names_from = province, values_from = n)
  
ncsc1 <- ncsc_wide %>% 
  select(-date) %>% 
  map(~ .x[.x > 0])

benford_analyses_ncsc1 <- map(ncsc1, benford, number.of.digits = 1)

table_ncsc1 <- make_statistics_table(benford_analyses_ncsc1)
print(table_ncsc1, n = Inf)

par(mfrow = c(16, 4), plt = c(.2, .97, .17, .9))
walk2(benford_analyses_ncsc1,
      names(benford_analyses_ncsc1),
      plot_digits_distribution)
```

### Specific comparisons

```{r fig.width = 2 * 4.24725, fig.height = 10 * 3.5 / 16}
ncsc_group1 <- ncsc_wide %>% 
  transmute(date, n = `Hai Duong` + `Quang Ninh`) %>% 
  filter(ymd(20210101) < date, date < ymd(20210401)) %>% 
  select(-date) %>% 
  map(~ .x[.x > 0])

ncsc_group2 <- ncsc_wide %>% 
  transmute(date, n = `Bac Giang` + `Bac Ninh`) %>% 
  filter(ymd(20210401) < date, date < ymd(20210801)) %>% 
  select(-date) %>% 
  map(~ .x[.x > 0])

ncsc_group3 <- ncsc_wide %>% 
  filter(ymd(20210501) < date, date < ymd(20211015)) %>% 
  select(-date) %>% 
  magrittr::extract(wave4mekong$province) %>% 
  rowSums() %>% 
  list()

ncsc_group4 <- ncsc_wide %>% 
  filter(date > ymd(20211015)) %>% 
  select(-date) %>% 
  rowSums() %>% 
  list()

ncsc_groups <- c(ncsc_group1, ncsc_group2, ncsc_group3, ncsc_group4) %>% 
  setNames(paste0("group", 1:4))

benford_analyses_ncsc_4groups <- map(ncsc_groups, benford, number.of.digits = 1)

make_statistics_table(benford_analyses_ncsc_4groups)

par(mfrow = c(1, 4), plt = c(.2, .97, .17, .9))
walk2(benford_analyses_ncsc_4groups,
      names(benford_analyses_ncsc_4groups),
      plot_digits_distribution)
```

